---
title: "The Big Five Advantage"
author: "Ella Trembanis"
description: "An analysis of score outcomes in the 2022 Eurovision final"
output:
  distill::distill_article:
    self_contained: false
---


## Introduction

The glitz, the glamor, the inexplicable banana-eating wolves - the Eurovision Song Contest is nothing if not a spectacle, and one I eagerly anticipate each year. The rules have always been, to say the least, a little fuzzy (who let Australia compete, anyway?), but one in particular strikes me as hostile to Eurovision's already dubious commitment to merit-based adjudication: France, Germany, Italy, Spain, and the United Kingdom - the so-called "Big Five" - automatically qualify for the finals, regardless of the quality of their performances. This guaranteed spot offers a leg up over other countries, which must survive earlier rounds of the competition to earn their place in the final, but a question remains: do the Big Five countries ultimately score better than their competitors?

In this project, I will test my hypothesis that Big Five countries do indeed out-score their competitors. I suspect that, since they compete in every final, Big Five countries will feel empowered to pour more resources into their respective contestants, and may have cultivated more audience loyalty over the years, compared to other countries which qualify on a more infrequent basis. A statistically significant, positive coefficient in my regression output would confirm my hypothesis. However, if my analysis reveals that there is no significant relationship between Big Five status and total score, or that such a relationship exists, but is negative (meaning Big Five countries score lower than their competitors), I will reject this hypothesis.

## Data

The data used in this project was compiled by user ODI6S on Kaggle, and can be found at https://www.kaggle.com/datasets/diamondsnake/eurovision-song-contest-data/data. My primary dataset, here titled "euro," consists of the point tallies (broken up into jury and televote scores) for all 25 countries that competed in the 2022 Eurovision Song Contest final. The study is cross-sectional, since it was compiled at one point in time, after the contest results were broadcast, and therefore does not capture any change over time. Additionally, the data collected is purely observational: my explanatory variable was not randomly assigned as part of a controlled manipulation of the contest. Since the causal assumptions have not been satisfied, my analysis will only apply to correlational relationships. Each row represents a single competing country.

Big Five status - which is to say, whether a country pre-qualified for the final round of competition - is the explanatory variable of interest here, and is coded as either "Big Five," for France, Germany, Italy, Spain, and the UK, or "Not Big Five," for all others. In order to incorporate other potential confounding variables in my regression analysis, I also incorporated another dataset from the same source, here titled "song," which includes other observational data from the 2022 contest, such as the gender of the performers and the language in which they sang. I will use each country's total score, the sum of their jury and televote points, as my response variable - below, I have produced a visualization of the distribution of this variable.

```{r setup}
library(tidyverse)
library(ggplot2)

# loading the dataset

euro <- read.csv("2022_jury_results.csv")

# creating a new variable

euro <- euro |>
  mutate(
    big_five = if_else(Contestant == "France" | 
                         Contestant == "Germany" |
                         Contestant == "Italy" | 
                         Contestant == "Spain" |
                         Contestant == "United Kingdom", 
                       "Big Five", "Not Big Five")
  )
```

```{r dependent var visualization}
# creating a visualization for the dependent variable (total score)

top_scores <- euro |>
  group_by(Contestant, big_five) |>
  summarize(
    score = mean(Total.score, na.rm = TRUE)
  )

ggplot(top_scores, mapping = aes(x = score, y = fct_reorder(Contestant, score))) +
  geom_col() +
  labs(
    title = "Distribution of 2022 Eurovision Final Scores",
    x = "Total Score",
    y = "Contestant"
  )
```

## Results

The following plot represents the average scores of the Big Five and non-Big Five groups in the 2022 final.

```{r barplot visualization}
# creating a barplot visualization

euro_bar <- euro |>
  group_by(big_five) |>
  summarize(
    avg_score = mean(Total.score, na.rm = TRUE)
  ) |>
  ggplot(mapping = aes(x = big_five, y = avg_score, fill = big_five)) +
  geom_col() +
  labs(
    title = "Average 2022 Eurovision Scores for Big Five and Non-Big Five Countries",
    x = "Big Five Status",
    y = "Average Total Score in 2022"
  ) +
  theme(legend.position = "none")

euro_bar
```

In this particular year, Big Five countries outperformed their competitors, receiving an average score of 243.2 to their competitors' 171.2. What remains to be seen, however, is whether this result is statistically significant.

The following regression models will attempt to shed some light on this question.

```{r single linear regression}
# creating a linear regression for the main variable

fit1 <- lm(Total.score ~ big_five, data = euro)

library(modelsummary)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "big_fiveNot Big Five" = "Not Big Five"
)

modelsummary(list(Regression = fit1),
             statistic = "p.value",
             fmt = fmt_decimal(digits = 2, pdigits = 3),
             coef_map = var_labels,
             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

This single regression confirms that not being in the Big Five was associated with a 72-point decrease in average total scores in 2022 (see the coefficient of "Not Big Five" in the above table). Yet the p-value of this coefficient is approximately 0.393. Since 0.393 is substantially greater than 0.05, we must accept the null hypothesis of no effect at a test level of $\alpha = 0.05$. In other words, this result is actually statistically indistinguishable from 0, and could very likely have resulted from random chance. Even if this regression had produced a significant result, it should not be interpreted causally, since Big Five status is not randomly assigned.

I will next produce a multiple regression model in order to assess the potential influence of confounders on this relationship.

```{r joining a dataset}
# loading a new dataset

song <- read.csv("song_data.csv")

song <- song |>
  filter(year == 2022, final_draw_position != "-") |>
  rename(Contestant = country) |>
  mutate(
    english = if_else(language == "English", "English", "Other"),
    dancers = if_else(backing_dancers > 0, "Backing Dancers", "No Backing Dancers"),
    singers = if_else(main_singers > 1, "Multiple Singers", "One Singer")
  )

# joining euro and song

euro_song <- left_join(euro, song, by = 'Contestant') |>
  select(Contestant, Total.score, Jury.score, Televoting.score, big_five, 
         final_draw_position, language, style, gender, main_singers, key, 
         BPM, energy, danceability, happiness, loudness, acousticness, 
         instrumentalness, liveness, speechiness, dancers, english,
         singers, backing_instruments, instrument_10, race)
```

```{r multiple regression}
# creating a multiple regression

fit2 <- lm(Total.score ~ big_five + gender + dancers + singers + english, data = euro_song)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "big_fiveNot Big Five" = "Not Big Five",
  "genderMale" = "Male",
  "genderMix" = "Mixed Gender",
  "dancersNo Backing Dancers" = "No Backing Dancers",
  "singersOne Singer" = "One Singer",
  "englishOther" = "Non-English Lyrics"
)

modelsummary(list(Regression = fit2),
             statistic = "p.value",
             fmt = fmt_decimal(digits = 2, pdigits = 3),
             coef_map = var_labels,
             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

This multiple regression indicates that not being in the Big Five is associated with a 141.31-point decrease in average total scores, where several other variables, including the gender of the performers, the number of back-up dancers and singers, and whether they sang non-English lyrics, are held fixed (see the coefficient of "Not Big Five" in the above table). The p-value for this coefficient - 0.136 - is less than the p-value from the previous, single regression, but it is still greater than 0.05, and therefore insignificant at a test level of $\alpha = 0.05$. The coefficients of the confounding variables are similarly insignificant. Even if this regression had produced a significant result, it should not be interpreted causally, since Big Five status is not randomly assigned.

## Conclusion

Based on the 2022 Eurovision final, there does not appear to be a statistically significant scoring advantage to being a Big Five member. Although my regressions produced a negative coefficient of "Not Big Five," confirming that Big Five countries outscored their competitors in 2022, these coefficients had high p-values, and so these results are not meaningfully different from 0, as they could very well have resulted from chance variation. In the absence of more compelling evidence, I therefore reject my initial hypothesis and accept the null hypothesis of no effect.

It should be noted that these conclusions are predicated entirely on one contest's results. Additional analysis of the performance of Big Five and non-Big Five countries over time could help clarify the full extent of this relationship. After all, it is also possible that I have committed a Type II error, and the relationship I hypothesized does exist, but is not perceivable at this level of analysis. Despite my incorporation of several potential confounders in the multiple regression, there are other factors at play which are not reflected in the dataset: for instance, current events may color voters' attitudes towards various countries - the war in Ukraine (a non-Big Five country) is one example, and could conceivably have contributed to that country's victory. Another avenue to extend this project would be to consider the differential impact of Big Five status on jury and televoting scores.
