---
title: "The Big Five Advantage"
author: "Ella Trembanis"
description: "An analysis of score outcomes in the 2022 Eurovision final"
output:
  distill::distill_article:
    self_contained: false
---


## Introduction

The glitz, the glamor, the inexplicable banana-eating wolves - the Eurovision Song Contest is nothing if not a spectacle, and one I eagerly anticipate each year. The rules have always been, to say the least, a little fuzzy (who let Australia compete, anyway?), but one in particular strikes me as hostile to Eurovision's already dubious commitment to merit-based adjudication: France, Germany, Italy, Spain, and the United Kingdom - the so-called "Big Five" - automatically qualify for the finals, regardless of the quality of their performances. This guaranteed spot offers a leg up over other countries, which must survive earlier rounds of the competition to earn their place in the final, but a question remains: do the Big Five countries ultimately score better than their competitors?

In this project, I will test my hypothesis that Big Five countries do indeed out-score their competitors. I suspect that, since they compete in every final, Big Five countries will feel empowered to pour more resources into their respective contestants, and may have cultivated more audience loyalty over the years, compared to other countries which qualify on a more infrequent basis. A statistically significant, positive coefficient in my regression output would confirm my hypothesis. However, if my analysis reveals that there is no significant relationship between Big Five status and total score, or that such a relationship exists, but is negative (meaning Big Five countries score lower than their competitors), I will reject this hypothesis.

## Data

The data used in this project was compiled by user ODI6S on Kaggle, and can be found at https://www.kaggle.com/datasets/diamondsnake/eurovision-song-contest-data/data. My primary dataset, here titled euro, consists of the point tallies (broken up into jury and televote scores) for all 25 countries that competed in the 2022 Eurovision Song Contest final. The study is cross-sectional, since it was compiled at one point in time, after the contest results were broadcast, and therefore does not capture any change over time. Additionally, the data collected is purely observational: my explanatory variable was not randomly assigned as part of a controlled manipulation of the contest. Since the causal assumptions have not been satisfied, my analysis will only apply to correlational relationships. Each row represents a single competing country.

Big Five status - which is to say, whether a country pre-qualified for the final round of competition - is the explanatory variable of interest here, and is coded as either "Big Five," for France, Germany, Italy, Spain, and the UK, or "Not Big Five," for all others. In order to incorporate other potential confounding variables in my regression analysis, I also incorporated another dataset from the same source, here titled song, which includes other observational data from the 2022 contest, such as the gender of the performers and the language in which they sang. I will use each country's total score, the sum of their jury and televote points, as my response variable - below, I have produced a visualization of the distribution of this variable.

```{r setup}
library(tidyverse)
library(ggplot2)

# loading the dataset

euro <- read.csv("2022_jury_results.csv")

# creating a new variable

euro <- euro |>
  mutate(
    big_five = if_else(Contestant == "France" | 
                         Contestant == "Germany" |
                         Contestant == "Italy" | 
                         Contestant == "Spain" |
                         Contestant == "United Kingdom", 
                       "Big Five", "Not Big Five")
  )
```

```{r dependent var visualization}
# creating a visualization for the dependent variable (total score)

top_scores <- euro |>
  group_by(Contestant, big_five) |>
  summarize(
    score = mean(Total.score, na.rm = TRUE)
  )

ggplot(top_scores, mapping = aes(x = score, y = fct_reorder(Contestant, score))) +
  geom_col() +
  labs(
    title = "Distribution of 2022 Eurovision Final Scores",
    x = "Total Score",
    y = "Contestant"
  )
```

## Results

```{r barplot visualization}
# creating a barplot visualization

euro_bar <- euro |>
  group_by(big_five) |>
  summarize(
    avg_score = mean(Total.score, na.rm = TRUE)
  ) |>
  ggplot(mapping = aes(x = big_five, y = avg_score, fill = big_five)) +
  geom_col() +
  labs(
    title = "Average 2022 Eurovision Scores for Big Five and Non-Big Five Countries",
    x = "Big Five Status",
    y = "Average Total Score in 2022"
  ) +
  theme(legend.position = "none")

euro_bar
```

```{r single linear regression}
# creating a linear regression for the main variable

fit1 <- lm(Total.score ~ big_five, data = euro)

library(modelsummary)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "big_fiveNot Big Five" = "Not Big Five"
)

modelsummary(list(Regression = fit1),
             statistic = "p.value",
             fmt = fmt_decimal(digits = 2, pdigits = 3),
             coef_map = var_labels,
             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

```{r joining a dataset}
# loading a new dataset

song <- read.csv("song_data.csv")

song <- song |>
  filter(year == 2022, final_draw_position != "-") |>
  rename(Contestant = country) |>
  mutate(
    english = if_else(language == "English", "English", "Other"),
    dancers = if_else(backing_dancers > 0, "Backing Dancers", "No Backing Dancers"),
    singers = if_else(main_singers > 1, "Multiple Singers", "One Singer")
  )

# joining euro and song

euro_song <- left_join(euro, song, by = 'Contestant') |>
  select(Contestant, Total.score, Jury.score, Televoting.score, big_five, 
         final_draw_position, language, style, gender, main_singers, key, 
         BPM, energy, danceability, happiness, loudness, acousticness, 
         instrumentalness, liveness, speechiness, dancers, english,
         singers, backing_instruments, instrument_10, race)
```

```{r multiple regression}
# creating a multiple regression

fit2 <- lm(Total.score ~ big_five + gender + dancers + singers + english, data = euro_song)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "big_fiveNot Big Five" = "Not Big Five",
  "genderMale" = "Male",
  "genderMix" = "Mixed Gender",
  "dancersNo Backing Dancers" = "No Backing Dancers",
  "singersOne Singer" = "One Singer",
  "englishOther" = "Non-English Lyrics"
)

modelsummary(list(Regression = fit2),
             statistic = "p.value",
             fmt = fmt_decimal(digits = 2, pdigits = 3),
             coef_map = var_labels,
             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


## Conclusion