---
title: "The Big Five Advantage"
author: "Ella Trembanis"
description: "An analysis of score outcomes in the 2022 Eurovision final"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

The glitz, the glamor, the inexplicable banana-eating wolves - the Eurovision Song Contest is nothing if not a spectacle, and one I eagerly anticipate each year. The rules have always been, to say the least, a little fuzzy (who let Australia compete, anyway?), but one in particular strikes me as contrary to Eurovision's already debatable commitment to merit-based adjudication: France, Germany, Italy, Spain, and the United Kingdom - the so-called "Big Five" - automatically qualify for the finals, regardless of the quality of their performances. This guaranteed spot offers a leg up over other countries, which must survive earlier rounds of the competition to earn their place in the final, but a question remains: do the Big Five countries ultimately score better than their competitors?

In this project, I will test my hypothesis that Big Five countries do indeed out-score their competitors. I suspect that, since they compete in every final, Big Five countries will feel empowered to pour more resources into their respective contestants, and may have cultivated more audience loyalty over the years, compared to other countries which qualify on a more infrequent basis. A statistically significant, positive coefficient in my regression output would confirm my hypothesis. However, if my analysis reveals that there is no significant relationship between Big Five status and total score, or that such a relationship exists, but is negative (meaning Big Five countries score lower than their competitors), I will reject this hypothesis.

## Data

```{r setup}
library(tidyverse)
library(ggplot2)

# loading the dataset

euro <- read.csv("2022_jury_results.csv")

# creating a new variable

euro <- euro |>
  mutate(
    big_five = if_else(Contestant == "France" | 
                         Contestant == "Germany" |
                         Contestant == "Italy" | 
                         Contestant == "Spain" |
                         Contestant == "United Kingdom", 
                       "Big Five", "Not Big Five")
  )
```

```{r dependent var visualization}
# creating a visualization for the dependent variable (total score)

top_scores <- euro |>
  group_by(Contestant, big_five) |>
  summarize(
    score = mean(Total.score, na.rm = TRUE)
  )

ggplot(top_scores, mapping = aes(x = score, y = fct_reorder(Contestant, score))) +
  geom_col() +
  labs(
    title = "Distribution of 2022 Eurovision Final Scores",
    x = "Total Score",
    y = "Contestant"
  )
```

## Results

```{r barplot visualization}
# creating a barplot visualization

euro_bar <- euro |>
  group_by(big_five) |>
  summarize(
    avg_score = mean(Total.score, na.rm = TRUE)
  ) |>
  ggplot(mapping = aes(x = big_five, y = avg_score, fill = big_five)) +
  geom_col() +
  labs(
    title = "Average 2022 Eurovision Scores for Big Five and Non-Big Five Countries",
    x = "Big Five Status",
    y = "Average Total Score in 2022"
  ) +
  theme(legend.position = "none")

euro_bar
```

```{r single linear regression}
# creating a linear regression for the main variable

fit1 <- lm(Total.score ~ big_five, data = euro)

library(modelsummary)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "big_fiveNot Big Five" = "Not Big Five"
)

modelsummary(list(Regression = fit1),
             statistic = "p.value",
             fmt = fmt_decimal(digits = 2, pdigits = 3),
             coef_map = var_labels,
             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

```{r joining a dataset}
# loading a new dataset

song <- read.csv("song_data.csv")

song <- song |>
  filter(year == 2022, final_draw_position != "-") |>
  rename(Contestant = country) |>
  mutate(
    english = if_else(language == "English", "English", "Other"),
    dancers = if_else(backing_dancers > 0, "Backing Dancers", "No Backing Dancers"),
    singers = if_else(main_singers > 1, "Multiple Singers", "One Singer")
  )

# joining euro and song

euro_song <- left_join(euro, song, by = 'Contestant') |>
  select(Contestant, Total.score, Jury.score, Televoting.score, big_five, 
         final_draw_position, language, style, gender, main_singers, key, 
         BPM, energy, danceability, happiness, loudness, acousticness, 
         instrumentalness, liveness, speechiness, dancers, english,
         singers, backing_instruments, instrument_10, race)
```

```{r multiple regression}
# creating a multiple regression

fit2 <- lm(Total.score ~ big_five + gender + dancers + singers + english, data = euro_song)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "big_fiveNot Big Five" = "Not Big Five",
  "genderMale" = "Male",
  "genderMix" = "Mixed Gender",
  "dancersNo Backing Dancers" = "No Backing Dancers",
  "singersOne Singer" = "One Singer",
  "englishOther" = "Non-English Lyrics"
)

modelsummary(list(Regression = fit2),
             statistic = "p.value",
             fmt = fmt_decimal(digits = 2, pdigits = 3),
             coef_map = var_labels,
             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


## Conclusion